Probing the inner kpc of massive galaxies with strong gravitational lensing

### 2015-01-05 ms.tex
more revisions for resubmission

Strong gravitational lensing is a powerful tool for probing the matter distribution in distant galaxies. Among other things, strong lenses have been used to constrain galaxy masses, density profiles, and abundance of dark matter subhalos \citep[e.g.,][]{dalal:02, Gavazzi:07, Bolton:08}. For double and quad image configurations, the theory of strong lensing predicts a third and a fifth image near the centers of lensing galaxies. 
\blue{Realistic galaxy density profiles produce central images that are almost always \emph{demagnified}\footnote{approximately as $(\Sigma/\Sigma_{crit})^{-2}$, where $\Sigma$ is the surface density and $\Sigma_{crit}$ is the critical surface density}, making their detection difficult \citep[e.g.,][]{Jackson:13}.}  
\red{Unlike the other lensed images, which are magnified, this image can be significantly \emph{demagnified}, making its detection difficult \citep[e.g.,][]{Jackson:13}.}  It is well-understood that the magnification of the central images is very sensitive to the matter distribution in the innermost regions of lens galaxies: very steep density profiles significantly demagnify the central images, whereas cored or shallow profiles render them brighter (Figure \ref{fig1}).

@@

The individual behavior and statistical properties of central images in lensed populations have been extensively studied \citep[e.g.,][]{Wallington:93, Evans:02, Keeton:03}. These studies show that central images could have a wide range of magnifications \citep[e.g.,][]{Keeton:03}. 
Observational searches for these images have found a number of candidates. Although observations
of central images in group and cluster lenses are not uncommon \citep[e.g.,][]{Inada:08}, only one secure detection of a central image of a galaxy-scale lens exists to date \citep{Winn:04}. To avoid the possibility of absorption in the lens, most searches have focused on lensed radio sources. The radio spectra of candidate central images have been used to distinguish them from potential faint emission from the lens galaxies \citep[e.g.,][]{Zhang:07}.

@@

As seen in this figure, larger stellar cores result in brighter images. 
In the absence of a massive black hole, when $\gamma_s=1.0$ and with a core size larger than 150 pc, such observations should be able to detect the central images with more than $5\sigma$ significance. For significantly shallower ($\gamma_s<0.8$) profiles and larger cores, the central image is \emph{magnified} and its detection should be trivial.  The magnification and flux of central images is also influenced by SMBHs.
SMBHs demagnify the images on scales comparable to the size of their Einstein radii ($\sim30$ mas for a $2\times10^8M_{\odot}$ SMBH).
\red{The dashed curves in Figure \ref{fig:3} show the suppression of the fluxes of central images in presence of a $2\times10^8M_{\odot}$ SMBH.}
\blue{A comparison of the thick black dashed curve to the solid black curve in Figure \ref{fig:3} shows the suppression of the fluxes of central images in presence of a $2\times10^8M_{\odot}$ SMBH.}

@@

We also find that the detection significance of the central images can significantly \emph{increase} for an extended source with a radius of 1 kpc compared to a point source. Since extended sources occupy a larger area of the source plane, which may cover regions with higher magnifications, the resulting flux can be larger than the flux of more compact sources. The solid blue curve in Figure \ref{fig:3} shows the increase in the flux of the demagnified image for a larger source (with a radius of 1 kpc). In contrast, the flux of a source with a radius of 250 pc can be severely supressed (dotted grey curve). 
As the source galaxies are undergoing prodigious star formation, these large source sizes are reasonable and in agreement with observations \citep{hezaveh:13b}.
More extended and intrinsically brighter emission lines such as C[II], the brightest molecular line in high redshift galaxies, may therefore be the most ideal targets for such observations.

>>>>

Strong gravitational lensing is a powerful tool for probing the matter distribution in distant galaxies. Among other things, strong lenses have been used to constrain galaxy masses, density profiles, and abundance of dark matter subhalos \citep[e.g.,][]{dalal:02, Gavazzi:07, Bolton:08}. For double and quad image configurations, the theory of strong lensing predicts a third and a fifth image near the centers of lensing galaxies. 
\blue{Realistic galaxy density profiles produce central images that are almost always \emph{demagnified}\footnote{approximately as $(\Sigma/\Sigma_{crit})^{-2}$, where $\Sigma$ is the surface density and $\Sigma_{crit}$ is the critical surface density}, making their detection difficult \citep[e.g.,][]{Jackson:13}.}  
%\red{Unlike the other lensed images, which are magnified, this image can be significantly \emph{demagnified}, making its detection difficult \citep[e.g.,][]{Jackson:13}.}
 It is well-understood that the magnification of the central images is very sensitive to the matter distribution in the innermost regions of lens galaxies: very steep density profiles significantly demagnify the central images, whereas cored or shallow profiles render them brighter (Figure \ref{fig1}).

@@

The individual behavior and statistical properties of central images in lensed populations have been extensively studied \citep[e.g.,][]{Wallington:93, Mao:01, Evans:02, Keeton:03}. These studies show that central images could have a wide range of magnifications \citep[e.g.,][]{Keeton:03}, \blue{and that cuspy profiles (or SMBHs) can suppress the flux of the central images \citep[e.g.,][]{Bowman:04,Rusin:05,Mao:12}.}
Observational searches for these images have found a number of candidates. Although observations
of central images in group and cluster lenses are not uncommon \citep[e.g.,][]{Inada:08}, only one secure detection of a central image of a galaxy-scale lens exists to date \citep{Winn:04}. To avoid the possibility of absorption in the lens, most searches have focused on lensed radio sources. The radio spectra of candidate central images have been used to distinguish them from potential faint emission from the lens galaxies \citep[e.g.,][]{Zhang:07}.

@@

As seen in this figure, larger stellar cores result in brighter images. 
In the absence of a massive black hole, when $\gamma_s=1.0$ and with a core size larger than 150 pc, such observations should be able to detect the central images with more than $5\sigma$ significance. For significantly shallower ($\gamma_s<0.8$) profiles and larger cores, the central image is \emph{magnified} and its detection should be trivial.  The magnification and flux of central images is also influenced by SMBHs.
SMBHs demagnify the images on scales comparable to the size of their Einstein radii ($\sim30$ mas for a $2\times10^8M_{\odot}$ SMBH).
%\red{The dashed curves in Figure \ref{fig:3} show the suppression of the fluxes of central images in presence of a $2\times10^8M_{\odot}$ SMBH.}
\blue{A comparison of the thick black dashed curve to the solid black curve in Figure \ref{fig:3} shows the suppression of the fluxes of central images in presence of a $2\times10^8M_{\odot}$ SMBH.}

@@

\begin{figure}
\begin{center}
\centering
\includegraphics[trim= 0 0 0 0, clip, width=0.48\textwidth]{figures/f_06.eps}
\centering
\end{center}
\vspace{-1mm}
\caption{\blue{The two top panels show the magnification cross-sections of two sources with different sizes (left: 100 pc, right 500 pc in radius) behind a lens containing a black hole with a mass of $1\times 10^8 M_{\odot}$. The magnification at the central regions of the left panel approaches zero and the central image does not form (the color scale has been clipped at $10^{-2.5}$). The lower panel shows the magnification at the center (when the source is perfectly behind the lens) as a function of source size).}
\label{figMu}}
\vspace{-1.5mm}
\end{figure}

We also find that the detection significance of the central images can significantly \emph{increase} for an extended source with a radius of 1 kpc compared to a point source. Since extended sources occupy a larger area of the source plane, which may cover regions with higher magnifications, the resulting flux can be larger than the flux of more compact sources. 
The solid blue curve in Figure \ref{fig:3} shows the increase in the flux of the demagnified image for a larger source (with a radius of 1 kpc). In contrast, the flux of a source with a radius of 250 pc can be severely supressed (dotted grey curve). 
\blue{This effect is illustrated in Figure \ref{figMu} where the effects of source size on magnification are shown. The two top panels of Figure \ref{figMu} show the magnification cross-section (flux magnification as a function of source position behind the lens) for two sources with different sizes (left: 100 pc, right: 500 pc in radius). For a compact source (top left panel) in the central regions the magnification approaches zero and the central image does not form. For more extended sources, however, even when the source is exactly behind the lens, the large extent of the source allows covering higher magnification regions, increasing the total flux magnification.
The bottom panel shows the magnification as a function of source size, for a source that is located on the lens axis.}
As the source galaxies are undergoing prodigious star formation, these large source sizes are reasonable and in agreement with observations \citep{hezaveh:13b}.
More extended and intrinsically brighter emission lines such as C[II], the brightest molecular line in high redshift galaxies, may therefore be the most ideal targets for such observations.
